---
title: "R-timeseries"
author: "Ryan Peek"
date: "Updated: `r format(Sys.Date())`"
output:
  github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("../"))
```

# Background Reading

These resources are very useful. Please read through them.

 - This is a great overview of times and dates in R from [Bonnie Dixon's talk at the Davis R-Users Group](http://www.noamross.net/blog/2014/2/10/using-times-and-dates-in-r---presentation-code.html)

 - [A very comprehensive and detailed look at using dates in R](https://www.stat.berkeley.edu/~s133/dates.html)

 - [The vignette on using lubridate](https://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html)

# Dates & Times in R


Importantly, there are **3** basic time classes in R:

 - Dates (just dates, i.e., 2012-02-10)
 - POSIXct ("**ct**" == calendar time, best class for dates with times)
 - POSIXlt ("**lt**" == local time, enables easy extraction of specific components of a time, however, remember that POXIXlt objects are lists)

So, for some brief examples/discussion about each of these classes, please read below. Unfortunately converting dates & times in R into formats that are computer readable can be frustrating, mainly because there is very little consistency. In particular, if you are importing things from Excel, keep in mind dates can get especially weird^[1], depending on the operating system you are working on, the format of your data, etc.

[^1]: For example Excel stores dates as a number representing days since 1900-Jan-0, plus a fractional portion of a 24 hour day (**serial-time**), but in OSX (Mac), it is 1904-Jan-0.
 
## Working with Dates

The `Date` class in R can easily be converted or operated on numerically, depending on the interest. Let's make a string of dates to use for our example:

 - `sample_dates_1 <- c("2016-02-01", "2016-03-21", "2016-10-05", "2017-01-01", "2017-02-18")` *notice we have dates across two years here*

So to translate these into the `Date` class in R, we need to do a few things...first notice our sample dates are in a very nice order (YYYY-MM-DD). This works well with the function `as.Date`: 

 - `as.Date(sample_dates_1)`

What happens with different orders...say MM-DD-YYYY? 

```{r sampledates, echo=T, eval=T}

# Some sample dates: 
sample_dates_2 <- c("02-01-2016", "03-21-2016", "10-05-2016", "01-01-2017", "02-18-2017")

as.Date(sample_dates_2) # well that doesn't work

```

Well, the reason is the computer expects one thing, but is getting something else. Remember, **write code you can read and your computer can understand**. So we need to give some more information here so this will interpret correctly.

```{r sampledates2, echo=T, eval=T}

# Some sample dates: 
sample_dates_2 <- c("02-01-2016", "03-21-2016", "10-05-2016", "01-01-2017", "02-18-2017")

as.Date(sample_dates_2, format = "%m-%d-%Y" ) # date code preceded by "%"

```

To see a list of the date-time format codes in R, check out this [page and table](https://www.stat.berkeley.edu/~s133/dates.html).

The nice thing is this method works well with pretty much any format, you just need to provide the associated codes and structure:

 - `as.Date("2016/01/01", format="%Y/%m/%d")`=`r as.Date("2016/01/01", format="%Y/%m/%d")`
 - `as.Date("05A21A2011", format="%mA%dA%Y")`=`r as.Date("05A21A2011", format="%mA%dA%Y")`
  - `as.Date("Feb 13, 2011", format="%b %d, %Y")`=`r as.Date("05A21A2011", format="%mA%dA%Y")`
 
### The `lubridate` package & Dates

The `lubridate` package will handle 90% of the date & datetime issues you need to deal with. It is fast, much easier to work with, and I recommend using it wherever possible. Do keep in mind sometimes you need to fall back on the base R functions (i.e., `as.Date()`). 

Same idea here, just simpler to type, read, write.

 - `lubridate::ymd("2016/01/01")`=`r lubridate::ymd("2016/01/01")`
 - `lubridate::ymd("2011-03-19")`=`r lubridate::ymd("2011-03-19")`
 - `lubridate::mdy("Feb 19, 2011")`=`r lubridate::mdy("Feb 19, 2011")`
 - `lubridate::dmy("22051997")`=`r lubridate::dmy("22051997")`
 
## Working with Datetimes & Timezones

Similar to `as.Date` there are two functions for each datetime class in R. `as.POSIXct` and `as.POSIXlt`. The arguments are the largely the same, but two important components to consider are a `strptime` function to help format the datetime (similar to `as.Date`), and a timezone argument `tz`, for adding the appropriate timezone.

Keep in mind the default datetime format is `YMD HM` or `YMD HMS`, so if you use that format you won't need to use the `strptime` option.

**as.POSIXct**

 - Example 1 (default doesn't need `strptime`): `as.POSIXct("2016-01-01 12:00", "America/Los_Angeles")` = `r as.POSIXct("2016-01-01 12:00", "America/Los_Angeles")`
 - Example 2 (24 hr time): `as.POSIXct(strptime("2016/04/05 14:47",format = "%Y/%m/%d %H:%M"), "America/Los_Angeles")` = `r as.POSIXct(strptime("2016/04/05 14:47",format = "%Y/%m/%d %H:%M"), "America/Los_Angeles")`
 - Example 3 (12 hr time): `as.POSIXct(strptime("2016/04/05 4:47 PM",format = "%Y/%m/%d %H:%M PM"), "America/Los_Angeles")` = `r as.POSIXct(strptime("2016/04/05 4:47 PM",format = "%Y/%m/%d %H:%M PM"), "America/Los_Angeles")`

### The `lubridate` package & datetimes

Same as before, now just add `_hms` or `_hm` to `ymd`. `lubridate` will default to the POSIXct format.

 - Example 1: `lubridate::ymd_hm("2016-01-01 12:00", tz="America/Los_Angeles")` = `r lubridate::ymd_hm("2016-01-01 12:00", tz="America/Los_Angeles")`
 - Example 2 (24 hr time): `lubridate::ymd_hm("2016/04/05 14:47", tz="America/Los_Angeles")` = `r lubridate::ymd_hm("2016/04/05 14:47", tz="America/Los_Angeles")`
 - Example 3 (12 hr time but converts to 24): `lubridate::ymd_hms("2016/04/05 4:47:21 PM", tz="America/Los_Angeles")` = `r lubridate::ymd_hms("2016/04/05 4:47:21 PM", tz="America/Los_Angeles")`

## Datetimes without Timezones (and `chron`)

The `chron` package may be helpful for these tasks, however, this may also be a suitable use of the POSIXlt class.

 - `chron::as.chron("2013-07-24 23:55:26")` = `r chron::as.chron("2013-07-24 23:55:26")`
 - `chron::as.chron("07/25/13 08:32:07", "%m/%d/%y %H:%M:%S")` = `r chron::as.chron("07/25/13 08:32:07", "%m/%d/%y %H:%M:%S")`

## Example & Challenge 1: Mauna Loa Meterological Data

So, now that we have a decent idea how to format these things, let's look at some real data, try to format and plot. Let's use the Mauna Loa meterological data, collected every minute for the year 2001. It's 459,769 observations for 9 different metrics of wind, humidity, barometric pressure, air temperature, and precipitation.

```{r mloa_data, eval=T, echo=T}

# let's load our climate data from our previous lesson:

load("data_output/mauna_loa_met_2001_minute.rda")

library(lubridate, warn.conflicts = F)
library(dplyr, warn.conflicts = F)

summary(mloa_2001)
names(mloa_2001)

```

One of the important components to consider is each of the datetime columns has been separated...so how do we get them into one column so we can format it as a datetime? The answer is the `paste` function.

 - `paste()` allows pasting text or vectors (& columns) by a given separator that you specify
 - `paste0()` is the same thing, but defaults to using a **`,`** as the separator.

```{r making datetime cols, echo=T, eval=T}

# we need to make a datetime column...let's use paste
mloa_2001$datetime <- paste0(mloa_2001$year,"-", mloa_2001$month, "-", mloa_2001$day," ", mloa_2001$hour24, ":", mloa_2001$min) # this makes a character column

# we can nest this within a lubridate function to convert directly to POSIXct
mloa_2001$datetime <- ymd_hm(paste0(mloa_2001$year,"-", mloa_2001$month, "-", mloa_2001$day," ", mloa_2001$hour24, ":", mloa_2001$min))

summary(mloa_2001) # notice a new column called "datetime"

```

```{r clean up data and plot March}

library(ggplot2)

# clean up the NA data (NA's are = -99 or -999 depending on data col)
mloa_2001 %>% 
  filter(!rel_humid == -99, 
         !temp_C_2m == -999,
         !windSpeed_m_s == -99.9,
         month==3) %>% 
  ggplot(.) + 
  geom_point(aes(x=datetime, y=rel_humid, color=temp_C_2m), alpha=0.5) + 
  viridis::scale_color_viridis() + ylim(0, 100)

```