---
title: "Spatial Data & R"
author: "Ryan Peek"
date: "Updated: `r format(Sys.Date())`"
output:
  github_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("../"))
```

## Libraries Required:

Let's look at the libraries we'll need for this lesson.
```{r libraries}

library(rgdal)  # for vector work; sp package will load with rgdal
library(raster) # for metadata/attributes- vectors or rasters
library(rasterVis)
library(leaflet) # interactive maps
library(readr) # reading stuff
library(dplyr) # doing stuff
library(viridis) # color palettes
```

## Coordinate Reference Systems (CRS)

How to “flatten” or project the spatial data in geographic space. Often called projection, this can be a tricky business if you don't know or have a specific projection/CRS for your data, or when trying to mix two different CRS at once.

Best is to use some standards, and transform all projections into the same CRS before you work/plot your data.

Check this for more:
 - CRS Projections: http://spatialreference.org/ref/epsg/
 - Choosing the Right Projection: https://source.opennews.org/articles/choosing-right-map-projection/

```{r CRS projections}

# set strings for UTMs and lats for commonly used projections (WGS84)
utms <- CRS("+init=epsg:32610") ## more detailed def using the EPSG code
lats <-CRS("+init=epsg:4326") # set the default for lat/longs, plays well with leaflet and google
```

## Get NEON Data

First we went and downloaded the [NEON Shapefile](https://ndownloader.figshare.com/files/3708751) data, which downloads as a *.zip* file. You'll need to double click on that file to unzip it (if on Mac), or try right clicking on the zip file and look for an unzip option. Make note of *where* the unzipped folder ends up. In my case I made a **`GIS`** folder and then unzipped it there. 

### RASTER Data

```{r raster}

DSM_HARV <- raster("data/GIS/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_dsmCrop.tif")

DSM_HARV # view the data

plot(DSM_HARV, 
     main="NEON Digital Surface Model\nHarvard Forest")

crs(DSM_HARV)

# assign crs to an object (class) to use for reprojection and other tasks
utmZ18 <- crs(DSM_HARV)
utmZ18

minValue(DSM_HARV)
maxValue(DSM_HARV)


hist(DSM_HARV,
     main="Distribution of Digital Surface Model Values\n Histogram Default: 100,000 pixels\n NEON Harvard Forest",
     xlab="DSM Elevation Value (m)",
     ylab="Frequency",
     col="wheat")

# notice warning...just telling us there were a lot of pixels not plotted
ncell(DSM_HARV)

# create histogram that includes with all pixel values in the raster
hist(DSM_HARV, 
     maxpixels=ncell(DSM_HARV),
     main="Distribution of DSM Values\n All Pixel Values Included\n NEON Harvard Forest Field Site",
     xlab="DSM Elevation Value (m)",
     ylab="Frequency",
     col="wheat4")
nlayers(DSM_HARV) # number of bands or layers
GDALinfo("data/GIS/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_dsmCrop.tif")

```

### VECTOR Data

```{r load shapefile}

aoiBoundary_HARV <- readOGR(dsn = "data/GIS/NEON-DS-Site-Layout-Files/HARV",
                            layer = "HarClip_UTMZ18")
crs(aoiBoundary_HARV)
extent(aoiBoundary_HARV)

lats # notice the differences
utms # notice the differences?

# access things in a spatial_BLANK_Dataframe with object_name@data
```


```{r quick plot}
plot(aoiBoundary_HARV, col="cyan1", border="black", lwd=3,
     main="AOI Boundary Plot")

```

#### CHALLENGE 1: Import More Data!

 - Get `HARV_roads` and `HARVtower_UTM18N` layers into R. Call the `Harv_roads` object **`lines_HARV`** and the `HARVtower_UTM18N` `**point_HARV**`.

```{r Challenge1, echo=F, eval=T}

aoiBoundary_HARV <- readOGR(dsn = "data/GIS/NEON-DS-Site-Layout-Files/HARV",
                            layer = "HarClip_UTMZ18")

lines_HARV <- readOGR(dsn = "data/GIS/NEON-DS-Site-Layout-Files/HARV",
                            layer = "HARV_roads")

point_HARV <- readOGR(dsn = "data/GIS/NEON-DS-Site-Layout-Files/HARV",
                      layer = "HARVtower_UTM18N")
# a way to turn off graphics device (start from scratch)
dev.off()

# Layering the plot

plot(aoiBoundary_HARV, col = "lightgreen", 
     main="NEON Harvard Forest\nField Site")
plot(lines_HARV, add = TRUE)

# use the pch element to adjust the symbology of the points
plot(point_HARV, add  = TRUE, 
     pch = 21, bg = "purple", cex=1.5)


# check extent of lines
extent(lines_HARV)
#xmin        : 730741.2 
#xmax        : 733295.5 
#ymin        : 4711942 
#ymax        : 4714260 

extent(aoiBoundary_HARV)
#xmin        : 732128 
#xmax        : 732251.1 
#ymin        : 4713209 
#ymax        : 4713359 

# change extent of boundaries using road layer
plot(aoiBoundary_HARV, col = "lightgreen", 
     main="NEON Harvard Forest\nField Site", xlim=c(730741.2,733295.5), ylim=c(4711942,4714260))

plot(lines_HARV, add = TRUE) # road

# use the pch element to adjust the symbology of the points
plot(point_HARV, add  = TRUE, 
     pch = 21, bg = "purple", cex=0.5)

```

#### CHALLENGE 2: Mix Vectors and Rasters

```{r Challenge2, echo=F, eval=T}

Crop_HARV <- raster("data/GIS/NEON-DS-Airborne-Remote-Sensing/HARV/CHM/HARV_chmCrop.tif")

aoiBoundary_HARV <- readOGR(dsn = "data/GIS/NEON-DS-Site-Layout-Files/HARV",
                            layer = "HarClip_UTMZ18")

lines_HARV <- readOGR(dsn = "data/GIS/NEON-DS-Site-Layout-Files/HARV",
                            layer = "HARV_roads")

point_HARV <- readOGR(dsn = "data/GIS/NEON-DS-Site-Layout-Files/HARV",
                      layer = "HARVtower_UTM18N")

# layer the plot together
plot(Crop_HARV, 
     main="Map of Harvard Forest Field Site\nStudy Area with Canopy Height Model (NEON)")

plot(aoiBoundary_HARV, col = "forestgreen", lwd=2,
     main="NEON Harvard Forest\nField Site", add=TRUE)

plot(lines_HARV, add = TRUE)

# use the pch element to adjust the symbology of the points
plot(point_HARV, add  = TRUE, pch = 21, bg = "purple", cex=1.2)

```


## Some other Tricks?

These are newish packages that you might find useful for GIS analysis, mapping ,etc. 

I'll talk about the `**leaflet**` package in particular, but there are many other spatial packages in R which may be useful.

```{r Extra_packages, echo=TRUE, eval=FALSE}

#devtools::install_github('walkerke/tigris')
library(tigris)

ua <- urban_areas(cb = TRUE) # Basic US urbanized areas data
us_geo <- states(class = "sf") # simplified feature of STATES

library(ggmap)
library(ggvis)
library(leaflet)
library(ggsn) # for scale bars/north arrows in ggplots

```

## Your Data!

Let's take a look at some of your stuff (thanks Ann and Tom!)

```{r holmes}
library(readr)

edna <- read_csv(file ="https://raw.githubusercontent.com/gge-ucd/wRangling_Seminar/master/data/GIS/eDNA_stations_19Feb2017_fixedc.csv")

edna <- read_csv("data/GIS/eDNA_stations_19Feb2017_fixedc.csv")

# Check how many NA's in spatial data
edna %>% filter(is.na(lat)) %>% tally # no NAs!

# make the data "spatial":
edna.SP <- SpatialPointsDataFrame(edna[, c("lon","lat")], edna)

# check the CRS
raster::crs(edna.SP) # should be NA
proj4string(edna.SP) <- lats # assign the CRS we want to use
raster::crs(edna.SP) # should be the full lats obj now

# Transform to UTMs (or whatever)
edna_utms.SP<-spTransform(edna.SP, utms)
raster::crs(edna_utms.SP) # should be the full UTMs now

# add UTMs as cols to dataframe?
edna_utms.SP@data$utm_y <- edna_utms.SP@coords[,1]
edna_utms.SP@data$utm_x <- edna_utms.SP@coords[,2]

```

```{r static plot}

library(maps)
library(mapdata)
suppressPackageStartupMessages(library(ggmap)) # ggplot functionality for maps
library(ggsn) # for scale bars/north arrows in ggplots

# see what extent is for our shape/points
extent(edna.SP)

# Use this in xlim/ylim
map("state",region=c('CA'), xlim = c(-122.3,-121.6), ylim=c(38,38.3))
map.axes()
map("county",region=c('CA'),boundary=FALSE,lty=3, add=TRUE)
plot(edna.SP, add=TRUE,pch=21, bg="#FDE725FF")
plot(edna.SP[edna.SP@data$tide_dir=="ebb",], add=TRUE,pch=16, cex=0.4, bg="#FDE725FF")


legend("topright", legend = c("eDNA-flood", "eDNA-ebb"), 
			 pt.bg = c("#FDE725FF"), col = "black", 
			 pch = c(21))
legend("topright", legend = c("eDNA-flood", "eDNA-ebb"), col = c(NA,"black"),
			 pch = c(NA, 16), bty="n", pt.cex = 0.4)


```


```{r map w background}

location=c(-121.95,38.1) # set the center of the map
map1 <- get_map(location=location,crop = F,
								color="bw",
								maptype="terrain",
								source="google",
								zoom=10)

sitemap <- ggmap(map1, extent = 'device') # use the BW terrain option
sitemap

library(ggrepel)
nicemap<-
	sitemap + 
	labs(x="Longitude (WGS84)", y="Latitude") + 
	geom_point(data=edna, aes(x=lon, y=lat), pch=21, size=2.4,
	           color='black',
	           fill="#2A788EFF", alpha=0.7) + 
	# geom_label_repel(data=edna, aes(x=lon, y=lat, label=Station), 
	# 					 nudge_y=0.01, size=3, fontface = "bold.italic", 
	# 					 label.r=unit(0.10, "lines"))+
	theme_bw()+
	theme(panel.border = element_rect(colour = "black", fill=NA, size=1))

nicemap

```

```{r leaflet map}

library(leaflet)

m <- leaflet() %>% addTiles() %>% 
  setView(lng = -121.95, lat = 38.1, zoom = 8) %>%  # zoom 5 for all CA 
  #addTiles(group = "OSM") %>%
  addProviderTiles("Stamen.TopOSMFeatures", group = "OSM Features") %>%
  addProviderTiles("Esri.WorldImagery", group = "ESRI Aerial") %>%
  addProviderTiles("Esri.WorldTopoMap", group = "ESRI Topo") %>%
	addProviderTiles("OpenStreetMap.BlackAndWhite", group = "B&W") %>%
	addProviderTiles("Stamen.TonerLite", group = "StamenT1") %>%
	addProviderTiles("Stamen.Watercolor", group = "StamenT2") %>%
	#addProviderTiles("Thunderforest.Landscape", group = "Topo") %>%
  hideGroup("OSM Features") %>% # busy layer so hide initially
  
	# other map layers:
	# Stamen.Toner
	# Stamen.TonerLite
	# Stamen.Watercolor
	
  # eDNA sites 
  addCircleMarkers(data=edna.SP, group="eDNA",
  								 popup=paste0("Station: ", edna.SP@data$Station, "<br>",
  								 						 "Month: ", edna.SP@data$collection_month, 
  								 						 "<br>","Tide: ",edna.SP@data$tide_dir,
  								 						 "<br>", "Trawl Type: ",edna.SP@data$trawl_type,
  								 						 "<br>", "Turbidity: ", edna.SP@data$turb),
                   stroke=TRUE, weight=0.6,radius=10,
                   fillOpacity = 0.8, color="black",
                   fillColor = "yellow") %>%

    # eDNA from dataframe  
  addCircleMarkers(data=edna.SP, group="eDNA_df",
                   lng = ~lon, lat = ~lat, 
                   radius = 6,
  								 popup=paste0("Station: ", edna$Station, "<br>",
  								 						 "Month: ", edna$collection_month, 
  								 						 "<br>","Tide: ",edna$tide_dir,
  								 						 "<br>", "Trawl Type: ",edna$trawl_type,
  								 						 "<br>", "Turbidity: ", edna$turb),
                   fillColor = ~ifelse(edna$tide_dir=="flood","darkblue", "red"),
  								 stroke=TRUE, weight=0.6,
                   fillOpacity = 0.8) %>%

  hideGroup("eDNA_df") %>%
		
# add a legend for points
  addLegend("bottomright", colors = c("darkblue", "red"),
            labels=c("Flood", "Ebb"),
            title = "Tidal Direction",
            opacity = 0.7) %>%

  # add controls for basemaps and data
  addLayersControl(
    baseGroups = c("OSM", "B&W", "StamenT1", "StamenT2", "ESRI Topo", "ESRI Aerial"),
    overlayGroups = c("eDNA","eDNA_df", "OSM Features"),
    options = layersControlOptions(collapsed = T))

m

# library(htmlwidgets)
# saveWidget(widget = m, file="eDNA_site_map.html", selfcontained = FALSE)

```

